A Dependently Typed Multi-Stage Calculus \ifnotanonymous\thanks{Supported by organization x.}\fi

### 2019-06-22 text/conclusion.tex
FInish revising Seciton 1.
Revise.

\section{Conclusions}

We proposed new multi-stage calculus \LMD, which can use dependent types.
\LMD enable finer-grained typing of multi-stage programming comparing to existing type systems.
This extention prevent some bugs relating to generated functions with multi-stage programming,
such as \verb|vadd| function for vectors of a fixed length.
We proved properties of \LMD including preservation, confluence, 
strong normalization for full reduction, and progress for staged reduction.
The technical difficulty of extention is how to deal with the symbol of CSP $\%$ in the equation rules.
We resolve this problem by ignoring $\%$ under some condition.

Making algorithmic typing and equality rules for \LMD is a future work.
They would be complicated but possible 
because we can realize them by defining a new weak head normal form which contains $\%$ erasing rule.

>>>>

\section{Conclusion \label{sec:conclusion}}

We have proposed a new multi-stage calculus \LMD with dependent types.
\LMD enables finer-grained typing of multi-stage programming comparing to existing type systems.
This extention prevents some bugs relating to generated functions with multi-stage programming,
such as \verb|vadd| function for vectors of a fixed length.
We proved properties of \LMD including preservation, confluence, 
strong normalization for full reduction, and progress for staged reduction.
The technical difficulty of extention is how to deal with the symbol of CSP $\%$ in the equation rules.
We resolve this problem by ignoring $\%$ under some condition.

Developing a typechecking algorithmic for \LMD is left for future work.
They would be complicated but possible 
because we can realize them by defining a new weak head normal form which contains $\%$ erasing rule.

### 2019-06-22 text/related-work.tex
FInish revising Seciton 1.
Revise.

\section{Related Work}

@@

There are many papers on dependent types and most of them are affected with the pioneering work of Per Martin-L\"{o}f.
$\lambda^\Pi$\cite{Meyer1986}, Calculus of Constructions\cite{coquand:inria-00076024}, 
and Harper, Honsell and Plotkin\cite{harper1993framework} are famous papers on dependent types.
In Advanced Topics in Types and Programming Languages\cite{attapl},
a dependent type system \LLF is designed with the basis of \cite{harper1993framework}.
We adopted \cite{attapl}-like dependent types for \LMD.

One can use dependent types in programming languages such as Idris\cite{brady2013idris} or
interactive theorem provers such as Coq\cite{09thecoq} which based on \cite{coquand:inria-00076024}.
\red{この段落要らない or 後ろとくっつけたほうが良いかも知れない?}

@@

Applications of dependent types for real problems were also researched.
In Xi and Pfenning\cite{Xi98}, they extended SML with restricted dependent types
and succeeded in reducing the bounds checking of arrays.
In Xi and Robert\cite{xi2001dependently}, they design a type system for an assembly language and
it is useful for spped up.
Xi also gave dead code elimination and loop unrolling as examples of dependent types application in \cite{xi1999dependent}.

\red{Besides,} multi-stage calculi also have a long history of research.
Davis revealed there is Curry-Howard correspondence between a multi-stage calculus and modal logic in $\lambda^\circ$\cite{davies1996temporal}.
However, $\lambda^\circ$ don't have operators for run and CSP.
\red{runはboldか? Besidesより適切な接続詞はないのか?}
Benaissa, et al. \cite{benaissa1999logical} study the relationship between multi-stage calculus and category theory or modal logic.
Taha and Sheard introduced run and CSP to a multi-stage calculus in \cite{MetaML}.
Additionally, Taha and Nielsen invented the concept of environment classifiers in $\lambda^\alpha$\cite{taha2003environment} and 
construct the type system for $\lambda^\alpha$, which can handle quoting, unquoting, run, and CSP.

In $\lambda^\TW$\cite{Tsukada}, Tsukada and Igarashi found modal logic which corresponds to a multi-stage calculus with environment classifiers and
show that run can be represented as application of $\varepsilon$ to transition abstractions.
In \LTP\cite{Hanada2014}, Hanada and Igarashi extended $\lambda^\TW$ with CSP and discuss code residualization 
which allows us to dump the quoted code into an external file.

Combination of multi-stage programming and another programming feature has been studied.
Oleg, Kameyama, and Sudo\cite{kiselyov2016refined} designed a type sysytem for multi-stage programming with mutable cells and
Oishi and Kameyama\cite{oishi2017staging} extended a multi-stage calculus so that it can handle control operators such as shift and reset.

MetaOCaml is a programming language with quoting, unquoting, run, and CSP.
Oleg gave many examples of application of MetaOCaml in \cite{8384206}, 
which include filtering in signal processing, matrix-vector product, and DSL compiler.

>>>>

\section{Related Work \label{sec:related-work}}

MetaOCaml is a programming language with quoting, unquoting, run, and
CSP.  Kiselyov give many applications of MetaOCaml in \cite{8384206},
which includes filtering in signal processing, matrix-vector product,
and DSL compiler.

Theoretical studies on multi-stage programming owe a lot to seminal
work by Davies and Pfenning~\cite{DaviesPfenning01JACM} and
Davies~\cite{davies1996temporal}, which found Curry-Howard
correspondence between multi-stage calculi and modal logic.  In
particular, Davies' $\lambda^\circ$~\cite{davies1996temporal} has been
a basis for multi-stage calculi with quasi-quotation.  $\lambda^\circ$
did not have operators for run and CSP; a few
studies~\cite{benaissa1999logical,MoggiTahaBenaissaSheard99ESOP}
enhanced and improved $\lambda^\circ$ towards the development of a
type-safe multi-stage calculus with quasi-quotation, run, and CSP,
which were proposed by Taha and Sheard as constructs for multi-stage
programming~\cite{MetaML}.
% Benaissa et al.~\cite{benaissa1999logical} study the relationship between a multi-stage calculus and category theory or modal logic.
% Taha and Sheard introduced run and CSP to a multi-stage calculus in 
Finally, Taha and Nielsen invented the concept of environment
classifiers~\cite{taha2003environment} and developed a typed calculus
$\lambda^\alpha$, which was equipped with all the features above in a
type sound manner and formed a basis of earlier versions of MetaOCaml.
A different approach to type-safe multi-stage programming with
different semantics for quasi-quotations has been studied by Kim, Yi,
and Calcagno~\cite{DBLP:conf/popl/KimYC06}.

Later, Tsukada and Igarashi~\cite{Tsukada} found correspondence
between a variant of \(\lambda^\alpha\) called $\lambda^\TW$
and modal logic and showed that run could be represented as a special
case of application of a transition abstraction ($\Lambda\alpha$) to
the empty sequence $\varepsilon$.  Hanada and
Igarashi~\cite{Hanada2014} developed \LTP as an extension
$\lambda^\TW$ with CSP.
% and discuss code residualization which
% allows us to dump the quoted code into an external file.

@@

There is much work on dependent types and most of them are affected by
the pioneering work by Martin-L\"{o}f~\cite{martin1973intuitionstic}.
Among many dependent type systems such as
$\lambda^\Pi$~\cite{Meyer1986}, Calculus of
Constructions~\cite{coquand:inria-00076024}, and Edinburgh
LF~\cite{harper1993framework}, we base our work on \LLF~\cite{attapl}
(which is quite close to $\lambda^\Pi$ and Edinburgh LF) due to its
simplicity.  It has been well known that dependent types are useful to
express detailed properties of data structure at the type level such
as the size of data structures~\cite{Xi98} and typed abstract syntax
trees~\cite{DBLP:conf/dsl/LeijenM99,DBLP:conf/popl/XiCC03}.  The
vector addition discussed in Section~\ref{sec:formal} is also such an
example.

@@

% Practical applications of dependent types have been also studied.
% One can use dependent types in programming languages such as Idris~\cite{brady2013idris} or
% interactive theorem provers such as Coq~\cite{09thecoq} based on \cite{coquand:inria-00076024}.
% In Xi and Pfenning~\cite{Xi98}, they extended SML with restricted dependent types
% and succeeded in reducing the bounds checking of arrays.
% In Xi and Harper~\cite{xi2001dependently}, they design a type system for an assembly language and
% it is useful for speed up.
% Xi also gave dead code elimination and loop unrolling as applications
%  of dependent types~\cite{xi1999dependent}.

% 関連研究との比較

Although there are studies on combinations of multi-stage programming and other programming features such as mutable cells~\cite{kiselyov2016refined},
control operators~\cite{KameyamaKiselyovShan09PEPM,oishi2017staging},
a combination with dependent types has been little studied.
One exception is Brady and Hammond~\cite{brady2006dependently},
who have discussed a
combination of dependently typeed programming with staging in the
style of MetaOCaml to implement a staged interpreter, which is
statically guaranteed to generate well typed code.  However, they
focused on concrete programming examples and there is no theoretical investigation
of the programming language they used.

% 他の方法としてどのようなものが考えられたか

% Although we define type equivalence of \LMD with composition of equivalence rules,
% there is another candiate to define,
% which gives reduction on types and compare the results of reduction such as~\cite{sorensen2006lectures}.
% This method is better than one of \LMD because equivalence rules become simple.
% However, we reject it because it cannot handle CSP flexibly.
